Using HMMs for pre-snap coverage detection

### 2024-12-09 submission.Rmd
New Delta

Choosing between man- and zone coverage is one of the most important strategic decisions a defensive coordinator has to take prior to each play in American football. While experienced coaches and quarterbacks can often identify these defensive strategies visually, the growing availability of tracking data presents another opportunity to infer the underlying tactic.
This project aims to leverage hidden Markov models (HMMs) to detect defensive strategies --- man or zone coverage --- based on pre-snap player movement data. 
By modeling hidden states that proxy the offensive player to be guarded, this approach provides a data-driven framework for unraveling the complexity of defensive patterns and enables real-time tactical insights for coaching and analysis.

@@

\textbf{UMSCHREIBEN} \textit{For all players and the ball carrier we use the features already contained in the tracking data, namely x- and y-coordinates, speed, acceleration, distance covered, orientation and direction. Except for the ball carrier we further compute the euclidean distance, x-distance and y-distance to the ball carrier. For defensive players only, we additionally compute the absolute difference of the defender’s direction and the angle of the shortest segment between the defender and the ball carrier. Subsequently, we order all players (in each frame) with respect to their euclidean distance to the ball carrier and standardize all features.
More details regarding data preprocessing can be found in the Appendix.}

@@

#### 1. Pre-motion analysis
We train a model to predict whether the defense plays a man- or zone coverage scheme. The 
model uses the previously described features, blablabla.

#### 2. HMM analysis
We model the movements of defensive players during the phase of pre-snap motion within a hidden Markov framework, in which the underlying states represent the offensive players to be guarded (see Franks et al., 2015 for a similar approach in basketball). In contrast to Groom et al., 2024, who enforce a state to proxy zone coverage during corner kicks in soccer, we cannot proceed similarly as the classical coverage zones in American football will only be covered by the defenders post-snap.

@@

A hidden Markov models consists of an observed time series $\{\boldsymbol{y}_t\}_{t=1}^T$ --- here, the y-coordinates of the defensive players and an unobserved first-order Markov chain $\{ g_t\}_{t=1}^T$, with $g_t \in \{1,\ldots,N\}$ which proxies the offensives players to be guarded at every time point $t$. The Markov chain is fully described by an initial distribution $\boldsymbol{\delta}=\bigl( \Pr(g_1=1), \ldots, \Pr(g_1=N) \bigr)$ and a transition probability matrix (t.p.m.) $\boldsymbol{\Gamma} = (\gamma_{ij}),$ with $\gamma_{ij} = \Pr(g_t = j| g_{t-1} = i), \ i,j = 1, \ldots, N$. The connection of both stochastic processes arises from the assumption that the distribution of the observations $\boldsymbol{y}_t$ are fully determined by the currently active state, i.e.\ 
\begin{equation*}
f(\boldsymbol{y}_t|g_1, \ldots, g_T, \boldsymbol{y}_1, \ldots, \boldsymbol{y}_{t-1},\boldsymbol{y}_{t+1},\ldots,\boldsymbol{y}_T) = f(\boldsymbol{y}_t|g_t).
\end{equation*}
In general, $f$ can be any density or probability mass function depending on the type of data. We are following the approaches of Franks et al. (2015) and Groom (2024) and opt for the Normal distribution.

@@

#### 3. Post-motion analysis
We re-train the pre-motion model to predict whether the defense plays a man- or zone coverage scheme, however, in this step, we incorporate the results from the HMM analysis.

#### 4. Motion evaluation
By comparing the predictive performance of our pre-motion model and our post-motion model we can determine the effectiveness of motion to detect the correct defensive scheme.

@@

All code for data preprocessing, model training, prediction and player evaluation can be found [here](https://github.com/janoleko/BDB-2025/).

>>>>

Choosing between man- and zone coverage is one of the most important strategic decisions a defensive coordinator has to take before each offensive play in American football. While experienced coaches and quarterbacks can often identify these defensive strategies visually, the growing availability of tracking data presents another opportunity to infer the tactics by the defence.
This project aims to leverage hidden Markov models (HMMs) to detect defensive strategies --- man or zone coverage --- based on pre-snap player movement data. 
By modeling hidden states that proxy the offensive player to be guarded, this approach provides a data-driven framework for unraveling the complexity of defensive patterns and enables real-time tactical insights for coaching and analysis.

@@

\textbf{UMSCHREIBEN} \textit{For all players and the ball carrier we use the features already contained in the tracking data, namely x- and y-coordinates, speed, acceleration, distance covered, orientation and direction. Except for the ball carrier we further compute the euclidean distance, x-distance and y-distance to the ball carrier. For defensive players only, we additionally compute the absolute difference of the defender’s direction and the angle of the shortest segment between the defender and the ball carrier. Subsequently, we order all players (in each frame) with respect to their euclidean distance to the ball carrier and standardize all features.
More details regarding data pre-processing can be found in the Appendix.}

@@

#### 1. Pre-motion analysis
We train a model to predict whether the defense plays a man- or zone coverage scheme. In particular, .....

The model uses the previously described features, blablabla.

#### 2. HMM analysis
We model the movements of defensive players during the phase of pre-snap motion within a hidden Markov framework, in which the underlying states represent the offensive players to be guarded (see Franks et al. 2015 for a similar approach in basketball). In contrast to Groom et al. (2024), who enforce a state to proxy zone coverage during corner kicks in soccer, we cannot proceed similarly as the classical coverage zones in American football will only be covered by the defenders post-snap.

@@

A hidden Markov models consists of an observed time series $\{\boldsymbol{y}_t\}_{t=1}^T$ --- here, the y-coordinates of the defensive players and an unobserved first-order Markov chain $\{ g_t\}_{t=1}^T$, with $g_t \in \{1,\ldots,N\}$ which proxies the offensives players to be guarded at every time point $t$. The Markov chain is fully described by an initial distribution $\boldsymbol{\delta}=\bigl( \Pr(g_1=1), \ldots, \Pr(g_1=N) \bigr)$ and a transition probability matrix (t.p.m.) $\boldsymbol{\Gamma} = (\gamma_{ij}),$ with $\gamma_{ij} = \Pr(g_t = j| g_{t-1} = i), \ i,j = 1, \ldots, N$. The connection of both stochastic processes arises from the assumption that the distribution of the observations $\boldsymbol{y}_t$ are fully determined by the currently active state, i.e.\ 
\begin{equation*}
f(\boldsymbol{y}_t|g_1, \ldots, g_T, \boldsymbol{y}_1, \ldots, \boldsymbol{y}_{t-1},\boldsymbol{y}_{t+1},\ldots,\boldsymbol{y}_T) = f(\boldsymbol{y}_t|g_t).
\end{equation*}
In general, $f$ can be any density or probability mass function depending on the type of data. Following the approaches of Franks et al. (2015) and Groom (2024), we opt for a Gaussian distribution.

@@

#### 3. Post-motion analysis
We re-train the pre-motion model to predict whether the defense plays a man- or zone coverage scheme, however, in this step, we incorporate results from the HMM analysis as further covariates.

#### 4. Motion evaluation
By comparing the predictive performance of our pre-motion model and our post-motion model we can determine the effectiveness of motion to detect the correct defensive scheme. Moreover, we can assess which teams predominantly apply pre-snap motions to increase the likelihood of correctly identifying the applied defensive strategy.

@@

All code for data pre-processing, model training, prediction and player evaluation can be found [here](https://github.com/janoleko/BDB-2025/).
